---
title: "CSSS/STAT 564 Lab Sessions #1"
author: Alex Ziyu Jiang
output: html_notebook
---

# Run a simple rstan example 

We first look at how to run a simple demo code in rstan. 

## The eigth schools example 

### Introduction

We will be looking at this classic example from Rubin(1981) and Gelman et. al.(2003). 

- Eight schools were tested on the effect on standardized test for coaching 
- Estimates and standard error of the treatment effect were calculated

|School|Estimated treatment effect|Std. error of treatment effect |
|------|---------|-----|
|A| 28| 15|
|B| 8 |10|
|C| -3| 16|
|D| 7 |11|
|E| -1| 9|
|F| 1 |11|
|G| 18| 10|
|H| 12| 18|

### The demo model

\begin{align*}
y_j &\stackrel{i.i.d}{\sim} N(\theta_j, \sigma_j^2), j = 1,...,J=8~~\text{(likelihood model)}  \\
\theta_j &\stackrel{i.i.d}{\sim} N(\mu , \tau), j = 1,...,J=8~~\text{(prior)} \\
p(\sigma_j,\mu, \tau) &\propto 1 ~~\text{(flat priors)}
\end{align*}

### breaking down the model

**observables**

- $y_j$: estimated treatment effects 
- $\sigma_j$: standard error of effect estimated 
- $J$: number of schools 

**unknown parameters**

- $\theta_j$: school treatment effects 
- $\mu$: population treatment effect
- $\tau$: standard deviation in school treatment effects

### key parts in a stan code

**required parts**

- data: data to be fed in the stan model
- parameters: unknown parameters in the model; goal of inference
- model: usually includes likelihood and prior

**optional parts**

- transformed parameters: preprocessing the parameter
- generated quantities: preprocessing the results 


```{r include=FALSE}
# load the package
library(rstan)  
# model compilation settings
options(mc.cores = parallel::detectCores()) # detect the number of CPU cores on the current host
rstan_options(auto_write = TRUE) # load the package
# set your working directory, basically where you expect your code outputs and your stan code to be
setwd("C:\\Users\\00000\\Documents\\GitHub\\CSSS-STAT-564\\lab1")
```

### the stan code (saved as '8schools.stan')

```{stan, eval = FALSE}
// not run
data {
  int<lower=0> J;         // number of schools 
  real y[J];              // estimated treatment effects
  real<lower=0> sigma[J]; // standard error of effect estimates 
}
parameters {
  real mu;                // population treatment effect
  real<lower=0> tau;      // standard deviation in treatment effects
}
model {
  theta ~ normal(mu, tau); // prior
  y ~ normal(theta, sigma); // likelihood
}
```

**important notes**

- vectorize your data: use <code>y[J]</code> to specify a vector of length J
- specify the range of your data: standard deviation/error has lower bound 0, use <code>lower=0</code>
- the second parameter in <code>normal()</code> is the standard deviation, not the variance

### run the model in rstudio

After coding the .stan file and saving it to the same directory of our R script, we can compile it in RStudio. 

```{r, message=FALSE, warning=FALSE}
# feed in the data (match the data part in your stan code)
schools_dat <- list(J = 8, 
                    y = c(28, 8, -3,  7, -1,1, 18, 12),
                    sigma = c(15, 10, 16, 11, 9, 11, 10, 18))
# run the stan code 
fit <- stan(file = "C:\\Users\\00000\\Documents\\GitHub\\CSSS-STAT-564\\lab1\\8schools.stan", data = schools_dat,refresh = 0)
```

We can call <code>stan_plot</code> to visualize the model output:

```{r}
stan_plot(fit, show_density = TRUE)
```

